\LARGE \bf Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks

### 2019-09-22 research/paper FG/paper.tex
Add FG paper format

\section{Introduction}

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing, the one that we are interested in, is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and it is important because of its applications to
\begin{itemize}
	\item call center conversations that improve service quality,
	\item speech translation systems that are more accurate at conveying emotions,
	\item and robotic pets that are more compassionate \cite{Koolagudi2012}.
\end{itemize}

@@

SER has historically been approached in two ways:
\begin{itemize}
	\item Feature engineering
	\item Machine learning (ML)
\end{itemize}

@@

\section{Methodology}

@@

\begin{table*}
	\centering
	\caption{Comparison of databases with their labeled emotions.}
	\label{dbEmoTable}
	\begin{tabular}{@{}lccccccccccccc@{}}
		\toprule
		Databases & \multicolumn{13}{c}{Emotions} \\ 
		\cmidrule{2-14}
		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
		\midrule
		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
		\bottomrule
	\end{tabular}%
\end{table*}

To build a more realistic SER system, we first need to obtain speech samples with their labeled emotion. We considered eight databases and chose four based on accessibility and based on the number of overlapping emotions. Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions. We chose four databases because it becomes more difficult to maintain consistency due to database variability. Databases can vary in
\begin{itemize}
	\item The set and number of labeled emotions
	\item The number of labels per sample
	\item The audio quality such as sampling rate and noise
	\item The spoken language
\end{itemize}
Given this variability, we chose the following four databases
\begin{enumerate}
	\item IEMOCAP \cite{busso_2008}
	\item TESS \cite{dupuis_2011}
	\item RAVDESS \cite{livingstone_2018}
	\item CREMA-D \cite{cao_2014}
\end{enumerate}

To control for the set and number of labeled emotions, we consider the following seven emotions for recognition
\begin{enumerate}
	\item Neutral
	\item Anger
	\item Disgust
	\item Fear
	\item Happy
	\item Sad
	\item Surprise
\end{enumerate}
We chose these seven emotions due to them being considered basic emotions by Ekman \cite{Ekman1992} and due to these seven being the most common among all databases.

@@

We combined these four databases into a combined database to train, validate, and test the neural network. The combined database is detailed in figure \ref{combinedDb}.

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{combined_db.png} 
	\caption{Proportions of each database, emotion, and label types in the combined database.}
	\label{combinedDb}
\end{figure*}

Each sample from a database went through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
\begin{enumerate}
	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a point in time.
	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were right-tail cropped and the extra information was discarded.
	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
	\item The sample is then converted into a log-Mel spectrogram using the STFT and Mel scale. The phase information was discarded as it does not appear to hold relevant information \cite{Kozakowski2017}.
	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a min-max scaling function.
\end{enumerate}

\begin{equation}
\label{stft}
\mathbf{STFT}\{x[n]\}(m,\omega) = \sum_{n=-\infty}^{\infty}x[n]w[n-m]e^{-j \omega n}
\end{equation}

The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function. We used a window size of 3072 with a 75\% overlap. This choice is based on the work of Zhao et al. \cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results.

\begin{equation}
\label{melScale}
m = 2595 log_{10}(1 + \frac{f}{700})
\end{equation}

The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels \cite{Oshaughnessy1990}. We set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins. The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms. The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution.

@@

\begin{figure}
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
	\caption{High level overview of the processing stages that a speech sample goes through.}
	\label{highLevelDataFlowDiagram}
\end{figure}

@@

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{neural_network_architecture.png}
	\caption{Architecture for the SER network including the input layer and eight processing layers.}
	\label{neuralNetworkArchitecture}
\end{figure*}

After all of the databases were processed as described in section \ref{Preprocessing}, the final combined database was fed into a neural network for training, validation, and testing. We constructed an eight layer neural network consisting of four convolutional layers and four dense layers. The network architecture is shown in figure \ref{neuralNetworkArchitecture}. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting \cite{Srivastava2014}, \cite{Ioffe2015}.
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

@@

\section{Results and Discussion}

After training the model, it was evaluated on the testing set to get the final accuracy. The final accuracy achieved was 57.64\%. Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix} and comparisons to the current literature is shown in table \ref{litacccompare}. We compare our results to the single-label case of SER as we could not find papers with results for the multi-label case.

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{confusion_matrix.png}
	\caption{Confusion matrices for each emotion.}
	\label{confusionMatrix}
\end{figure*}

@@

\section{Conclusion and Future Work}

@@

\begin{acks}
This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.
\end{acks}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8
%
% Slightly modified by X. Baró for FG2020
%

%\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4
                                                          % paper
\usepackage{FG2020}

%\FGfinalcopy % *** Uncomment this line for the final submission

\IEEEoverridecommandlockouts                              % This command is only
                                                          % needed if you want to
                                                          % use the \thanks command
\overrideIEEEmargins
% See the \addtolength command later in the file to balance the column lengths
% on the last page of the document

% The following packages can be found on http:\\www.ctan.org
%\usepackage{graphics} % for pdf, bitmapped graphics files
\usepackage{booktabs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage{hyperref}
\usepackage{cite}

%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed

\def\FGPaperID{****} % *** Enter the FG2020 Paper ID here

\title{\LARGE \bf
Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks
}

%use this in case of a single affiliation
\author{\parbox{16cm}{\centering
    {\large Brian Pho, Thomas Truong, and Svetlana Yanushkevich}\\
    {\normalsize
    Biometric Technologies Laboratory, Department of Electrical and Computer Engineering, University of Calgary, Alberta, Canada\\}}
}

%use this in case of several affiliations
%\author{\parbox{16cm}{\centering
%    {\large Huibert Kwakernaak$^1$ and Pradeep Misra$^2$}\\
%    {\normalsize
%    $^1$ Faculty of Electrical Engineering, Mathematics and Computer Science, University of Twente, Enschede, The Netherlands\\
%    $^2$ Department of Electrical Engineering, Wright State University, Dayton, USA}}
%    \thanks{This work was not supported by any organization}% <-this % stops a space
%}

\begin{document}

\ifFGfinal
\thispagestyle{empty}
\pagestyle{empty}
\else
\author{Anonymous FG2020 submission\\ Paper ID \FGPaperID \\}
\pagestyle{plain}
\fi
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

Current speech emotion recognition systems often overlook the multi-label data that comes with databases. In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech. We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 57.64\%. This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{INTRODUCTION}

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing, the one that we are interested in, is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and it is important because of its applications to: call center conversations that improve service quality, speech translation systems that are more accurate at conveying emotions, and robotic pets that are more compassionate \cite{Koolagudi2012}.

@@

SER has historically been approached in two ways: feature engineering and machine learning (ML).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{METHODOLOGY}

@@

%\begin{table*}
%	\centering
%	\caption{Comparison of databases with their labeled emotions.}
%	\label{dbEmoTable}
%	\begin{tabular}{@{}lccccccccccccc@{}}
%		\toprule
%		Databases & \multicolumn{13}{c}{Emotions} \\ 
%		\cmidrule{2-14}
%		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
%		\midrule
%		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
%		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
%		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
%		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
%		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
%		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
%		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
%		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
%		\bottomrule
%	\end{tabular}%
%\end{table*}

To build a more realistic SER system, we first collect speech samples with their labeled emotion. We considered eight databases and chose four based on accessibility and based on the number of overlapping emotions. We chose four databases because it becomes more difficult to maintain consistency due to database variability. Databases can vary in
\begin{itemize}
	\item The set and number of labeled emotions
	\item The number of labels per sample
	\item The audio quality such as sampling rate and noise
	\item The spoken language
\end{itemize}
Given this variability, we chose the following four databases: IEMOCAP \cite{busso_2008}, TESS \cite{dupuis_2011}, RAVDESS \cite{livingstone_2018}, and CREMA-D \cite{cao_2014}.

To control for the set and number of labeled emotions, we consider the following seven emotions for recognition: neutral, anger, disgust, fear, happy, sad, and surprise. We chose these seven emotions due to them being considered basic emotions by Ekman \cite{Ekman1992} and due to these seven being the most common among all databases.

@@

We combined these four databases into a combined database to train, validate, and test the neural network. Each sample from a database went through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
\begin{enumerate}
	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a point in time.
	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were right-tail cropped and the extra information was discarded.
	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
	\item The sample is then converted into a log-Mel spectrogram using the STFT and Mel scale. The phase information was discarded as it does not appear to hold relevant information \cite{Kozakowski2017}.
	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a min-max scaling function.
\end{enumerate}

For the STFT, we used a window size of 3072 with a 75\% overlap. This choice is based on the work of Zhao et al. \cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results.

For the Mel scale, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins. The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms. The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution.

@@

%\begin{figure}
%	\centering
%	\hspace{6mm}
%	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
%	\caption{High level overview of the processing stages that a speech sample goes through.}
%	\label{highLevelDataFlowDiagram}
%\end{figure}

@@

%\begin{figure*}
%	\centering
%	\hspace{6mm}
%	\includegraphics[width=\textwidth]{neural_network_architecture.png}
%	\caption{Architecture for the SER network including the input layer and eight processing layers.}
%	\label{neuralNetworkArchitecture}
%\end{figure*}

After all of the databases were processed as described in section \ref{Preprocessing}, the final combined database was fed into a neural network for training, validation, and testing. We constructed an eight layer neural network consisting of four convolutional layers and four dense layers. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting \cite{Srivastava2014}, \cite{Ioffe2015}.
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

@@

\addtolength{\textheight}{-3cm}   % This command serves to balance the column lengths
                                  % on the last page of the document manually. It shortens
                                  % the textheight of the last page by a suitable amount.
                                  % This command does not take effect until the next page
                                  % so it should come on the page before the last. Make
                                  % sure that you do not shorten the textheight too much.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{RESULTS AND DISCUSSION}

After training the model, it was evaluated on the testing set to get the final accuracy. The final accuracy achieved was 57.64\%. Comparisons to the current literature is shown in table \ref{litacccompare}. We compare our results to the single-label case of SER as we could not find papers with results for the multi-label case.

\begin{figure}
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{confusion_matrix.png}
	\caption{Confusion matrices for each emotion.}
	\label{confusionMatrix}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CONCLUSION AND FUTURE WORK}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ACKNOWLEDGMENTS}

This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,references}

\end{document}
